# Introduction


```{r load_common_objects, echo=FALSE}
  
 source("common.R")

```

As part of the training we receive from Professor Rafael A. Irizarry while taking the Data Science course (PH125.9x) of HarvardX, we commissioned to create a Machine Learning project based on the MovieLens data and write a report with our results. With this challenge, we remembered all the course teachings and utilized our newly acquired knowledge in R and Data Science.

The goal of this project is to create a recommendation algorithm for movies. This specific recommendation algorithm is going to predict and suggest movies that users might be interested in watching. The (training and test) data set was already prepared for us and it is a subset of the larger MovieLens data set[^1_introduction-1]. 

```{r setting_minimum_RMSE,echo=FALSE}

minimum_RMSE <- 0.8649 # as set by the project's guidelines
  
```

```{r saving_minimum_RMSE, echo=FALSE}

saveRDS(minimum_RMSE, file = paste0(path_of_files, "objects/minimum_RMSE.rds"))

```

The baseline for this project is to minimize the **Root Mean Squared Error (or RMSE)** of predicted ratings for pairs of user/movie, below the threshold of `r minimum_RMSE`.

```{r showing_minimum_RMSE,echo=TRUE,eval=FALSE}

minimum_RMSE <- 0.8649 # as set by the project's guidelines
  
```

[^1_introduction-1]: <https://grouplens.org/datasets/movielens/10m/>

Recommendation algorithms is not something new and it is already used for many different fields like search engines (the magical thing that fills in your sentence while you type what you are searching for), shopping suggestions in eShops (e.g. Amazon.com, eBay.com, etc), movie suggestions (e.g. Netflix, Disney+, etc).

To learn more about recommender systems you can read the following article in Wikepedia[^1_introduction-2] or in NVIDIA's web site[^1_introduction-3].

[^1_introduction-2]: <https://en.wikipedia.org/wiki/Recommender_system>

[^1_introduction-3]: <https://www.nvidia.com/en-us/glossary/recommendation-system/>


## How this report is organized 

This report is organised as follows. 

In Chapter 2, we describe the data set and prepare it for our analysis.  We also add some helper functions that we will use later.  

Chapter 3 is showing some visualizations of our data that will help us understand what the features and characteristics of our data are. 

In Chapter 4 we are setting the baseline that the success of our models will be measured against.  

Chapter 5 proposes, trains and predicts results with the help of three Machine Learning models (from the caret package).  All these models were selected with our goal in mind and all of them are (also) used for regression (some of them are also used for classification tasks).

Chapter 6 attacks the problem from a different angle by utilizing a specialized Matrix Factorization model (from the recommender package).  This model is designed especially for recommendation problems like this.

In Chapter 7 we describe our findings and propose possible ways that may improve the RMSE even more.

Lastly, in Chapter 8 we mention the technical characteristics of the underlying hardware and software that was used for carrying out our analysis.

All the R code has been extensively commented (see # ...) in order for the reader to be able to understand what we are doing on each point and why we are doing it.

In case you have any suggestions or recommendations on how we can improve this report, please leave us a comment on the github.
