# Appendix


## System Information


For the shake of reproducability we include below the specifications of the machine (computer) that was used for carrying out our experiments and the successful completion of this project.  

This project was carried out on a machine with the following hardware and software characteristics.  In case you try to reproduce the results and your R Studio is crashing, please make sure that your machine has these minimum hardware requirements (especially regarding the Total R.A.M Memory available).

```{r echo=FALSE}

library(parallel) 

system_info <- Sys.info()
number_of_cores <- detectCores()

print(paste0("OS: ", system_info["sysname"], " version ", system_info["version"], " (release: ", system_info["release"], ")"))
print(paste0("Virtual Machine: ", system_info["machine"]))
print(paste0("CPU: 2 x Intel Xeon E5-2697 v2 @ 2.70 GHz (from which we used 1 processor with ", number_of_cores, " cores)"))
print(paste0("Total R.A.M memory: 40 GB"))

```

## Session Info

In the section below the version information of R, the OS and attached (or loaded packages) are showing.

```{r echo=FALSE}

sessionInfo()

```